---
title: "Class 19: Pertussis Mini Project"
author: "Alejandro Ostos A17978684"
format: pdf
toc: TRUE
---

## Background

Pertussis is a bacterial lung infection also known as Whooping cough. Let's begin by examining CDC reported case numbers in the US. 

```{r}
cdc <- data.frame(
                          Year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,
                                   2019L,2020L,2021L,2022L,2023L,2024L),
  No..Reported.Pertussis.Cases = c(107473,
                                   164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617,6124,
                                   2116,3044,7063,22538)
)
head(cdc)
```

Plot of cases per year for Pertussis in the US

```{r}
library(ggplot2)

ggplot(cdc, aes(x = Year, y = No..Reported.Pertussis.Cases)) +
  geom_line(size = .60) +
  geom_point() +
  labs(title = "Reported Pertussis Cases in the US (1922-2023)",
       x = "Year",
       y = "Number of Reported Cases") 
```

Add some major milestone timepoints into our plot:

```{r}
ggplot(cdc, aes(x = Year, y = No..Reported.Pertussis.Cases)) +
  geom_line(size = .60) +
  geom_point() +
  labs(title = "Reported Pertussis Cases in the US (1922-2023)",
       x = "Year",
       y = "Number of Reported Cases") +
  geom_vline(xintercept = 1947, linetype="dashed", color = "red") +
  geom_vline(xintercept = 1996, linetype="dashed", color = "blue") +
  geom_vline(xintercept = 2020, linetype="dashed", color = "darkgreen") +
  annotate("text", x = 1947, y = 45000, label = "Routine vaccination", angle=90, vjust=-0.5, color="red") +
  annotate("text", x = 1996, y = 65000, label = "Switch to aP vaccine", angle=90, vjust=-0.5, color="blue") +
  annotate("text", x = 2020, y = 60000, label = "Lockdown", angle=90, vjust=-0.5, color="darkgreen")
```

The full introduction of mandatory wP (whole-cell) Pertussis immunization in the mid 1940s lead to a dramatic reduction in case numbers (from over 200,000 to 100s).

The switch to the aP (newer acellular) formalization precedes a gradual increase in case numbers over the next two decades, peaking in 2012 with over 48,000 reported cases.

Finally, the recent COVID-19 lockdowns and social distancing measures appear to have led to a sharp drop in Pertussis cases in 2020 and 2021, with a slight rebound in 2024.

## The CMI-PB project

The mission of the CMI-PB project is to develop a next-generation Pertussis vaccine that provides longer-lasting immunity and better protection against circulating strains.

They make their data available via JSON format API endpoints- basically the databse tables in a key:value format. To read this we can use the `jsonlite` package.

```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE)
head(subject)
```

> Q. How many "subjects"/individuals are in this dataset?

```{r}
nrow(subject)
```


> Q. How many wP and aP subjects are there?

```{r}
table(subject$infancy_vac)
```

>Q. What is the breakdown by "biological_sex" and "race"?

```{r}
table(subject$biological_sex)
```

```{r}
table(subject$race)
```

```{r}
table(subject$race, subject$biological_sex)
```

This breakdown is not particularly representative of the US population - this is a serious caveat when interpreting the results of this study. However, it is still the largest sample of it's type every assembled.

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE)
ab_titer <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = TRUE)
head(specimen)
```

```{r}
head(ab_titer)
```

We need to "join" or link these tables with the `subject` table so we can begin to analyze this data and know who a given Ab sample was colected for and when.

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)
head(meta)
```

Now let's join the `ab_titer` table to  `meta` to get the actual Ab measurements.

```{r}
ab_data <- inner_join(meta, ab_titer)
head(ab_data)
```
> Q. How many Ab measurements are there in total?

```{r}
nrow(ab_data)
```

> Q. How mnany different isotypes are there in the `ab_data` table?

```{r}
unique(ab_data$isotype)
```

> Q. How many different antigens are there in the `ab_data` table?

```{r}
unique(ab_data$antigen)
```

```{r}
ggplot(ab_data)+
  aes(MFI, antigen)+
  geom_boxplot()
```

IgG is crucial for long-term immunity against Pertussis, so let's filter to just IgG measurements.

```{r}
ab_data |>
  filter(isotype == "IgG") -> igg
```

```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen)+
  geom_boxplot(position=position_dodge(width=0.8))+
  labs(title="IgG Antibody Levels by Antigen and Infant Vaccine Type",
       x="IgG MFI",
       y="Antigen")
```

We can "facet" our plot by wP vs aP infant vaccine type to see if there are any differences.

```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen, colour = infancy_vac)+
  geom_boxplot(position=position_dodge(width=0.8))+
  labs(title="IgG Antibody Levels by Antigen and Infant Vaccine Type",
       x="IgG MFI",
       y="Antigen") + 
  facet_wrap(~infancy_vac)
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  xlim(0,75) +
  theme_bw()
```


More advanced analysis digging into individual antigen responses over time:

```{r}
filter(igg, antigen=="PT") |>
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = T) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

```{r}
  filter(igg,  antigen == "PT", dataset == "2021_dataset") |>
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

This graph shows the IgG response to Pertussis Toxin (PT) over time in individuals who received either wP or aP infant vaccines and were boosted with Tdap as adults. The peak in 14 days post-boost happens due to the rapid activation of memory B cells generated from prior vaccination. 

There are lots of cool things to explore in this dataset and we need coding and biology knowledge to do it effectively - i.e. you guys!!!



