---
title: "Class 8: Breast Cancer Mini Project"
author: Alejandro Ostos PID A17978684
format: gfm
toc: true
---
# Class 8

## Background
The goal of this mini-project is for you to explore a complete analysis using the unsupervised learning techniques covered in class. You’ll extend what you’ve learned by combining PCA as a preprocessing step to clustering using data that consist of measurements of cell nuclei of human breast masses. This expands on our RNA-Seq analysis from last day.

# Section 1
## Data Import

```{r}
fna.data <- "https://bioboot.github.io/bimm143_S20/class-material/WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names = 1)
```

```{r}
head(wisc.df)
```

```{r}
wisc.data <- wisc.df[,-1]
```

```{r}
diagnosis <- wisc.df[,1]
head(diagnosis)
```

> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.df)
```

There are 569 observations in the dataset.

> Q2. How many of the observations have a malignant diagnosis?

```{r}
sum(diagnosis == "M")
```

There are 212 malignant diagnosis.

```{r}
sum(diagnosis == "B")
```

There are 357 benignant diagnosis.

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
mean_cols <- grep("_mean$", colnames(wisc.df), value = TRUE)
length(mean_cols)
```

There are 10 variabes in the data that have the _mean suffix. 

# Section 2

The `prcomp()` function to perform a PCA has a `scale = FALSE` default. In general we want to set this to true, because our information has different units and scaling uniforms our data. Scaling will let us work with the data and do an accurate PCA.

```{r}
colMeans(wisc.data)
apply(wisc.data,2,sd)
```

```{r}
wisc.pr <- prcomp(wisc.data, scale = T, center = T)
summary(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

44.27% of the original variance is captured by the first component.

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

We need 3 components to describe at least 70% of the original variance data, because PC3 has a cumulative proportion of 72.363%. 

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

We need 7 components to describe at least 90% of the original variance data, because PC7 has a cumulative proportion of 91.01%

```{r}
biplot(wisc.pr)
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

This plot is very difficult to understand because all the information is clustered together and we are not even able to read the labels. There is a lot of information in one single plot, and differentiating between the variables is impossible. 


```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
    aes(PC1, PC2, col = diagnosis) +
    geom_point()
```

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
ggplot(wisc.pr$x[, 1 & 3]) +
    aes(PC1, PC3, col = diagnosis) +
    geom_point()
```

The first plot spreads out more to the negative of PC2, while the second plot spreads out to the positive of PC3. Also, the second plot has more overlap between the two groups. 

```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)

```

```{r}
pve <- pr.var / sum(pr.var)

plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

```{r}

library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation["concave.points_mean",1]
```
The component of the loading vector for `concave.points_mean` is -0.2609.

> Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

```{r}
summary(wisc.pr)
```

The minimum number of principal components required to explain 80% of the variance of the data is 5 or PC5 because its cumulative proportion is 0.84734 or 84.734%

# Section 3
## Hierarchical clustering

```{r}
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled, method = "euclidean")
wisc.hclust <- hclust(data.dist, method = "complete")
```

> Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h = 18, col="red", lty=2)
```


```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)
table(wisc.hclust.clusters, diagnosis)
```

> Q12. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 9)
table(wisc.hclust.clusters, diagnosis)
```

> Q13. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

# Section 5
## Combining Methods (PCA and Clustering)

Clustering the original data was not very productive. The PCA results looked promising. Here we combine these methods by clustering from our PCA results. In other words "clustering" in PC space...

```{r}
## Take the first 3 PCs 
dist.pc <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(dist.pc, method = "ward.D2")
```

View the tree...
```{r}
plot(wisc.pr.hclust)
abline(h = 70, col = "red")
```

To get our clustering membership vector (i.e. our main clustering result) we "cut" the tree at a desired height or to yield a desired number of "k" groups. 

```{r}
grps <- cutree(wisc.pr.hclust, k = 2)
table(grps, diagnosis)
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
```

```{r}
g <- as.factor(grps)
levels(g)
```

```{r}
g <- relevel(g,2)
levels(g)
```

```{r}
plot(wisc.pr$x[,1:2], col=g)
```

```{r}
wisc.pr.dist <- dist(wisc.pr$x[, 1:7])  
wisc.pr.hclust <- hclust(wisc.pr.dist, method = "ward.D2")
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
table(wisc.pr.hclust.clusters, diagnosis)
```
> Q15. How well does the newly created model with four clusters separate out the two diagnoses?

It does a good job, but there are still some dots that are not correctly classified. Cluster 1 should have all Malignant diagnosis, while cluster 2 should have all the B diagnosis. The 28 B and 24 M are false positives that should be arranged in the other cluster.

```{r}
table(wisc.hclust.clusters, diagnosis)
table(grps, diagnosis)
```
> Q16. How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.

It does a worse job because it does not separate the clusters according to the diagnosis accurately. And it performed clustering on 9 groups which is not what we are looking for. Also, I did not do the optional work, thus I do not have `wisc.km$cluster. The grps table does a good job with few false positives. It created 2 clusters, 1 having mostly M while the other having mostly B.

> Q17. Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

Sensitivity: TP/(TP+FN)
Specificity: TN/(TN+FN)

wisc.pr.hclust: Sensitivity ≈ 88.7%, Specificity ≈ 92.2% -> Best Sensitivity
wisc.hclustL: Sensitivity ≈ 81.6%, Specificity ≈ 96.6% -> Best Specificity
grps: Sensitivity ≈ 84.5% and Specificity ≈ 93.3%

# Section 7
## Prediction

We can use our PCA model for prediction with new input patient samples.

```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```
```{r}
plot(wisc.pr$x[,1:2], col=g)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Q18. Which of these new patients should we prioritize for follow up based on your results?

Patient 2 should be prioritized because it falls under the malignant diagnosis on the PCA. Patient 1 falls under the benign cases, thus it is more probable that his diagnosis is benign. 
