---
title: "Class 5: Data Viz with ggplot"
author: "Alejandro Ostos (PID: A17978684)"
format: gfm
---

Today we are exploring the **ggplot** package and how to make nice figures in R.

There are lots of ways to make figures and plot in R. These include:

- so called "base" R
- and add on packages like **ggplot2**

Here is a simple "base" R plot.
```{r}
head(cars)
```

We can simply pass to the `plot()` function
```{r}
plot(cars)
```

> Key-point: Base R is quick but not so nice and simple looking in some folk eyes. 

Let's see how we can plot this with **ggplot2**

1st I need to install this add on package. For this we use the `install.packages()` function. **WE DO THIS IN THE CONSOLE, NOT OUR REPORT** This is a one time thing. 

2nd We need to load the package with the `library()` 
```{r}
library(ggplot2)
ggplot(cars)
```
Every ggplot is composed of at least three layers:
- **data** (i.e. a data.frame with the things you want to plot)
- aesthetics **aes()** that map the columns of data to your plot features (i.e. aesthetics)
- geoms like **geom_point()** that srt how the plot aappears

```{r}
ggplot(cars) +
  aes( x = speed, y = dist) +
  geom_point()
```

> For simple "canned" graphs base R is quicker but as things get more custom and elaborate then ggplot wins out... 


Let's add more layers to our ggplot

Add a line showing a relationship between x and y
Add a Title
Add custom axis labels "Speed (MPH)" and "Distance (ft)"
Change the theme...

```{r}
ggplot(cars) +
  aes( x = speed, y = dist) +
  geom_point(color = "lightgray") +
  geom_smooth(method = lm, se = F, color = "orange") +
  labs(title = "Silly plot of Speed vs Stopping distance",
       x = "Speed (MPH)",
       y = "Stopping Distance (ft)") +
  theme_dark()
```

## Going Further

Read some gene expression data

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```


> Q1. How many genes are in this wee dataset?

```{r}
nrow(genes)
ncol(genes)

```

> Q2. How many "up" regulated genes are there?

```{r}
upreg <- genes$State == "up"
sum(upreg)
```

A useful function for counting up occurances of things in a vector is the `table()` function.


```{r}
table(genes$State)
```


```{r}
p <- ggplot(genes) + 
    aes(x = Condition1, y = Condition2, colour = State) +
    geom_point()
p
```


```{r}
p <- p + scale_colour_manual( values=c("red","lightgray","darkgreen") ) +
  theme_bw() +
  labs( title = "Gene Expression Changes Upon Drug Treatment",
        x = "Control (No Drug)",
        y = "No Treatment")

p
```


##More Plotting

Read the gapminder dataset


```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

Let's have a wee peak

```{r}
head(gapminder, 3)
```

```{r}
tail(gapminder, 3)
```


>Q4. How many different country values are in this dataset?

```{r}
length(table(gapminder$country))
```

> Q5. How many different continent values are in this dataset

```{r}
length(table(gapminder$continent))
```


```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col = continent) +
  geom_point() +
  labs(title = "GDP per capita vs Life Expectancy") 
  
```


```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col = continent, label = country) +
  geom_point() +
  labs(title = "GDP per capita vs Life Expectancy") +
  geom_text()
```


I can use **ggrepel** package to make more sensible labels here. 

```{r}
library(ggrepel)
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col = continent, label = country) +
  geom_point() +
  labs(title = "GDP per capita vs Life Expectancy") +
  geom_text_repel() 


```


I want a seperate pannel per continent

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col = continent, label = country) +
  geom_point() +
  labs(title = "GDP per capita vs Life Expectancy") +
  facet_wrap(~continent)
```


The main advantages of ggplot over base R plot are:

1. **Layered Grammar of Graphics**: ggplot uses a consistent, layered approach where you build plots by adding layers (data, aesthetics, geoms, themes) with the + operator. This makes complex plots easier to construct and modify [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

2. **Declarative Syntax**: You specify *what* you want to show (data mappings, aesthetics) rather than *how* to draw each element. This leads to more readable and maintainable code [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

3. **Beautiful Defaults**: ggplot produces publication-quality figures with attractive default styles, reducing the need for manual tweaking [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

4. **Customization and Extensibility**: It’s easier to add custom layers, annotations, and themes, and to map variables to visual features like color, size, and shape [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

5. **Consistency Across Plot Types**: The same grammar applies to scatter plots, bar plots, box plots, etc., making it easier to learn and use for different visualizations [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

Base R plots are quick for simple visualizations but can be fiddly and time-consuming to refine for publication-quality output [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

Which of these advantages do you think would be most useful for your work?