---
title: "Class 07: Machine Learning 1"
author: "Alejandro Ostos PID: A17978684"
format: pdf
---

Today we will explore some fundamental machine learning methods including clustering and dimensionality reduction.

## K-means clustering

To see how this works let's first makeup some data to cluster where we know what the answer should be. We can use the `rnorm()` function to help here:

```{r}
hist(rnorm(500, 5))
```
```{r}
x <- c(rnorm(30, -3), rnorm(30, 3))
y <- rev(x)
```

```{r}
x <- cbind(x, y)
```

```{r}
plot(x)
```


The function for K-means clustering in "base" R is `kmeans()`

```{r}
k <- kmeans(x, centers = 2)
k
```

To get the results of the returned list object we can use the dollar `$` syntax.

> Q. How many dots are in each cluster?

```{r}
k$size
```

> Q. What ‘component’ of your result object details
 - cluster assignment/membership?
 - cluster center?
 
```{r}
k$cluster
```

```{r}
k$centers
```

> Q. Make a clustering results figure of the data  colored by cluster membership and show membership and show cluster centers. 

```{r}
plot(x, col = c("red", "darkgreen"))
```

```{r}
plot(x, col = 4 + k$cluster, pch = 16)
points(k$centers, col = "orange", pch = 18, cex = 2)
```

K-means clustering is very popular as it is very fast and relatively straight forward: it takes numeric data as input and returns the cluster membership vector etc. 

The "issue" is we tell `kmeans()` how many clusters we want!

> Q. Run k-means again and cluster into four groups/clusters and plot the results like we did above?

```{r}
k4 <- kmeans(x, centers = 4)
k4
```
```{r}
plot(x, col = 4 + k4$cluster, pch = 16)
points(k4$centers, col = "orange", pch = 18, cex = 2)
```

brute-force
```{r}
k1 <- kmeans(x, center = 1)
k2 <- kmeans(x, center = 2)
k3 <- kmeans(x, center = 3)
k4 <- kmeans(x, center = 4)
k5 <- kmeans(x, center = 5)
```

```{r}
z <- c(k1$tot.withinss, k2$tot.withinss,
k3$tot.withinss,
k4$tot.withinss,
k5$tot.withinss)

plot(z, typ = "b" )
```


```{r}
n <- NULL
for (i in 1:5){
  n <- c(n, kmeans(x, centers = i)$tot.withinss)
}
plot(n, typ = "b")
```

## Hierarchical Clustering

The main "base" R function for Hierarchical Clustering is called `hclust()`. Here we can't just input our data we need to first calculate a distance matrix (e.g. `dist()`) for our data and use this as input to `hclust()`.

```{r}
d <- dist(x)
hc <- hclust(d)
hc
```
There is a plot method for hclust results lets try it. 

```{r}
plot(hc)
abline(h = 8, col = "red")
```

To get our cluster "membership" vector (i.e. our main clustering result) we can "cut" the tree at a given heigh or at a height that yields a given "k" groups.

```{r}
cutree(hc, h = 8)
```
```{r}
grps <- cutree(hc, k = 2)
```

> Q. Plot the data with our hclust result coloring

```{r}
plot(x, col=grps)
```


# Principal Component Analysis (PCA)

## PCA of UK food data

Import food data from an online CSV file:

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
head(x)
```
```{r}
rownames(x) <- x [,1]
x <- x[,-1]
x
```

```{r}
x <- read.csv(url, row.names = 1)
x
```

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```

```{r}
pairs(x, col=rainbow(10), pch=16)
```

> Main point: It can be difficult to spot major trends and patterns even in relatively small multivariate datasets (here we only have 17 dimensions, typically we have 1000s). 

## PCA to the rescue

The main function in "base" R for PCA is called `prcomp()`.

I will take the transpose of our food data so the "foods" are in the columns:

```{r}
pca <- prcomp(t(x))
summary(pca)
```
```{r}
pca$x
cols <- c("orange", "red", "blue", "darkgreen")
plot(pca$x[,1], pca$x[,2], col = cols, pch = 16)
```

```{r}
library(ggplot2)
```

```{r}
ggplot(pca$x) +
  aes(PC1, PC2) +
  geom_point(col=cols)
```

```{r}
pca$rotation
```

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```

PCA looks super useful and we will come back to describe this further next day :-)