---
title: "Class 10: Halloween Mini Project"
author: "Alejandro Ostos (PID A17978684)"
format: pdf
toc: TRUE
---

As it is nearly Halloween and the half way point in the quarter let's do a mini project to help us figure out the best candy!

Our data comes from the 538 website and is available as a CSV file>:

```{r}
library(readr)
candy <-  read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv", row.names=1)
head(candy)

```

```{r}
library(flextable)
flextable::flextable(head(candy))
```

>Q1. How many different candy types are in this dataset?

```{r, message=FALSE}
library(tidyverse)
candy %>%
  nrow()
```

>Q2. How many fruity candy types are in the dataset?

```{r}
candy$fruity %>%
  sum()
```

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favourite candy is Trolli so...
```{r}
candy["Trolli Sour Bites", ]$winpercent
```


>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```
 ## Quick overview of the data set
 
```{r}
library(skimr)
skimr::skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Winpercent is on a different scale than the rest of variables, because it does not range between 0 and 1. Although, winpercent is a percentage which means that it also ranges between 0 and 1. The variables that only have 0 or 1s are binaries, because they only represent a virtual yes or no. 

>Q7. What do you think a zero and one represent for the candy$chocolate column?

A zero would represent no chocolate, while a 1 would represent chocolate. 

>Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy)+
  aes(winpercent)+
  geom_histogram(bins = 20)
```

>Q9. Is the distribution of winpercent values symmetrica?

The distribution is not symmetrical, it is skewed right.

```{r}
ggplot(candy)+
  aes(winpercent)+
  geom_density()
```


>Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

```{r}
summary(candy$winpercent)
```

The center of the distribution is below 50 because the median is 47.83. This happens even if the mean is above the 50%. 

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# 1. Find all chocolate candy in the dataset
# 2. Find their winpercent values
# 3. Calculate the mean of these values

# 4-6. Do the same for fruity candy
# 7. Compare mean winpercents for chocolate vs. fruity

choc.inds <- candy$chocolate == 1
choc.win <- candy[choc.inds, ]$winpercent
mean(choc.win)

frt.inds <- candy$fruity == 1
frt.win <- candy[frt.inds, ]$winpercent
mean(frt.win)


```

Chocolate candy is on average ranked higher than fruity candy. 

>Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, frt.win)
```
This means are statistically significant. 

>Q13. What are the five least liked candy types in this set?

```{r}
candy |>
  arrange(winpercent)|>
  head(5)
```

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy |>
  arrange(-winpercent)|>
  head(5)
```


>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```


>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy)+
  aes(x=winpercent, y=reorder(rownames(candy), winpercent))+
  geom_col()
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```

>Q17. What is the worst ranked chocolate candy?

Sixlets are the worst ranked chocolate candy.

>Q18. What is the best ranked fruity candy?

Starburst are the best ranked fruity candy. 

## Winpercent and Pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label= rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
candy |>
  mutate(bang_for_buck = winpercent / pricepercent) |>
  arrange(desc(bang_for_buck)) |>
  slice(1)
```


>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
top5_expensive <- candy |>
  arrange(desc(pricepercent)) |>
  head(5)

top5_expensive

least_popular <- top5_expensive |>
  arrange(winpercent) |>
  slice(1)

least_popular
```


## Exploring the correlation structure

```{r}
candy <- candy[,-1]
cij <- cor(candy)
cij
```

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

```{r}
which(cij == min(cij), arr.ind = TRUE)
```
The two most anti-correlated variables are pluribus and bar. This because they are represented in red and are the most negative values in the plot

>Q23. Similarly, what two variables are most positively correlated?

```{r}
cij_no_diag <- cij
diag(cij_no_diag) <- NA
which(cij_no_diag == max(cij_no_diag, na.rm = TRUE), arr.ind = TRUE)
```
The most correlated variables are bar and nougat because they are represented in blue meaning that they are the most positive values. 

## Principal Component Analysis

The function to use is called `prcomp()` with an optional `scale=T/F` argument. 

```{r}
candy <- candy[,-1]
pca <- prcomp(candy, scale = T)
summary(pca)
```

Our main PCA result figure

```{r}
ggplot(pca$x) + 
  aes(PC1, PC2) +
  geom_point(col = my_cols)+
  geom_text_repel(col = my_cols, label = rownames(candy))

```

We should also examine the variable "loadings" or contributions of the original variables to the new PCs
```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```


```{r}
library(plotly)
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```
Interactive plots that can be zoomed on and "brushed" over can be made with the **plotly** package. It's output is interactive and will not render to PDF :(.
```{r}
#plotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

The variables that are picked up strongly by PC1 in the positive direction are pluribus and hard. It kind of makes sense to me because it groups hard and pluribus candies together, and in a certain way when imagining a bag with a lot of candy I imagine mass production candies that are normally hard as well. 


